Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-02-11 jso.qmd
update title
edits on sec 5

---
title: New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial/Zhuhai Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimises a criterion function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optimum has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. They are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour. The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  tandf-pdf:
    keep-tex: true
    fontsize: 12pt
    linestretch: 2
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
editor_options: 
  chunk_output_type: console
header-includes:
   - \usepackage{algorithm}
   - \usepackage{float} 
   - \usepackage{amsmath}
   - \usepackage{amsthm}
   - \usepackage{amssymb}
   - \theoremstyle{plain}
   - \newtheorem{defn}{\protect\definitionname}
   - \newtheorem{prop}{\protect\propositionname}
   - \providecommand{\definitionname}{Definition}
   - \providecommand{\propositionname}{Proposition}
nocite: |
   @DLMF
---

@@

This section describes the details of two simulation studies: one
compares the JSO performance with an existing optimiser, Creeping Random
Search (CRS) [@RJ-2021-105; @laa_using_2020] to explore JSO's behaviour
under different data dimensions. The second simulation first examines
the JSO performance under different hyper-parameters combinations
(number of jellyfish and the maximum number of tries), and then studies
the effect of index properties (smoothness and squintability), along
with JSO hyper-parameters, and data dimension, on the JSO performance
through a logistic regression.

@@

The JSO performance is investigated both in comparison to the existing
optimizer, CRS, in a pipe-finding problem using the holes index in the
6, 8, 10, and 12-dimensional space. Fifty simulations are conducted in
four data dimensions ($d = 6, 8, 10, 12$) with each optimiser. JSO uses
100 jellyfish with a maximum of 100 tries, while the CRS allows a
maximum of 1000 samples at each iteration before the algorithm
terminates. The different numbers account for the multiple paths of JSO
to enable fairer comparison with CRS. The results of the simulation are
collected using the data structure proposed in @RJ-2021-105 for
assessing JSO, where the design parameters are stored along with index
value, projection basis, random seed, and computation time.

@@

To examine JSO's performance across various hyper-parameter
combinations, the pipe-finding problem using the holes index is repeated
for fifty times in each hyper-parameter combination: 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts. The success rate is
calculated for each combination.

To assess JSO's performance across various PP problems, two different
data shapes, pipe and a sine wave, are investigated in 6D and 8D spaces
using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. Again,
JSO is run 50 times for each scenario to calculate the success rate for
each PP problem.

>>>>

---
title: Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial/Zhuhai Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimises a criterion function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optimum has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. They are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour. The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  tandf-pdf:
    keep-tex: true
    fontsize: 12pt
    linestretch: 2
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
editor_options: 
  chunk_output_type: console
header-includes:
   - \usepackage{algorithm}
   - \usepackage{float} 
   - \usepackage{amsmath}
   - \usepackage{amsthm}
   - \usepackage{amssymb}
   - \theoremstyle{plain}
   - \newtheorem{defn}{\protect\definitionname}
   - \newtheorem{prop}{\protect\propositionname}
   - \providecommand{\definitionname}{Definition}
   - \providecommand{\propositionname}{Proposition}
nocite: |
   @DLMF
---

@@

This section describes the details of two simulation studies. The first simulation
compares the JSO performance with an existing optimiser, Creeping Random
Search (CRS) [@RJ-2021-105; @laa_using_2020] to explore JSO's behaviour
under different data dimensions. The second simulation examines various factors affecting the performance of JSO.
The JSO performance is evaluated under different hyper-parameters combinations
(number of jellyfish and the maximum number of tries). The effect of index properties (smoothness and squintability) along
with JSO hyper-parameters and data dimension variations is captured and studied using logistic regression.

@@

The JSO performance is compared to the existing optimizer, CRS, in a pipe-finding problem using the holes index in different dimensional spaces. Fifty simulations are conducted in four data dimensions ($d = 6, 8, 10, 12$) with each optimiser. JSO uses
100 jellyfish with a maximum of 100 tries, while the CRS allows a
maximum of 1000 samples at each iteration before the algorithm
terminates. The different numbers account for the multiple paths of JSO
to enable fairer comparison with CRS. The results of the simulation are
collected using the data structure proposed in @RJ-2021-105 for
assessing JSO, where the design parameters are stored along with index
value, projection basis, random seed, and computation time.

@@

To examine the performance of JSO across various hyper-parameter
combinations, the pipe-finding problem using the holes index is repeated
for fifty times in each hyper-parameter combination: 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts. The success rate is
calculated for each combination.

In addition, to assess the performance of JSO across various PP problems, two different
data shapes, pipe and a sine wave, are studied in 4D, 6D and 8D spaces
using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. Again,
JSO is run 50 times for each scenario to calculate the success rate for
each PP problem.
